---
title: Spotify Popularity Data using Bayesian Hierarchical Models
author: "Israh Imam"
date: "3/30/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(bayesrules)
library(tidyverse)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(broom.mixed)
library(forcats)

library(kableExtra)
library(formattable)
library(data.table)
library(plotly)
library(wordcloud)
library(tidyselect)
library(tm)
library(lubridate)

library(psych)
library(DAAG)
library(highcharter)
library(knitr)
library(DT)
library(corrplot)
library(leaps)

```

# Results & Analysis
#### Data Exploration

```{r Data Import, include=FALSE}

# Import data
all_tracks <- read.csv("all_tracks_updated.csv", header = TRUE, sep = ",")

```

```{r Grouping, include=FALSE} 

all_tracks %>% group_by(Genre = genre) %>%
  summarise(`Track Count` = n()) %>% arrange(desc(`Track Count`)) %>%
    knitr::kable()

```

```{r Genre Density Plots, echo=FALSE, message=FALSE}

all_tracks <- all_tracks %>% arrange(desc(popularity))
top_songs <- all_tracks[c(1:200), ]

drop <- "Other"
all_tracks_wo_other <- top_songs[!(top_songs$genre == drop), ]

songs_data <- names(all_tracks_wo_other)[c(7:14, 16, 18)]
songs <- all_tracks_wo_other %>% 
  select(c('genre', songs_data)) %>%
  pivot_longer(cols = songs_data)

songs %>%
  ggplot(aes(x = value)) +
  geom_density(aes(color = genre)) +
  facet_wrap(~name, ncol = 4, scales = 'free') +
  labs(title = 'Genres & their Characteristics', x = '', y = '') +
  theme(axis.text.x = element_text(angle = 50, hjust = 1), axis.text.y = element_blank())

```


\hfill\break

```{r Characteristic Correlation, echo=FALSE}

corr_plot <- all_tracks_wo_other %>%
  select(popularity, acoustic, danceability, energy, instrumentalness, liveness,
         loudness, speechiness, tempo, key, mode, valence, length_mins)

corrplot(cor(corr_plot),
         method = 'color',
         type = 'upper',
         order = 'hclust')

```


```{r Popular Song Features, echo=FALSE, message=FALSE, warning=FALSE}

# Drop unecessary columns
all_tracks_drop <- all_tracks_wo_other[!duplicated(all_tracks_wo_other$song_title),]
all_tracks_drop <- all_tracks_wo_other[c(-2, -4, -5, -19, -20)]
colnames(all_tracks_drop)[1] <- "song_title"

track_features <- names(all_tracks_drop)[c(4:11, 13:14, 15)]

all_tracks <- all_tracks_drop %>% 
  arrange(desc(popularity)) %>%
  head(n = 500) %>%
  pivot_longer(cols = track_features)

all_tracks %>% 
  ggplot(aes(x = name, y = value)) +
  geom_jitter(aes(color = genre)) +
  facet_wrap(~name, ncol = 3, scales = 'free') +
  labs(title = 'Feature Patterns', x = '', y = '') +
  theme(axis.text.y = element_blank())

```

\hfill\break

```{r Examining Characteristics, echo=FALSE, message=TRUE, warning=FALSE}

top_songs <-
  all_tracks_drop %>%
  select(song_title, artist, genre, popularity, acoustic, danceability, energy,
         loudness, speechiness, tempo, key, liveness) %>%
  group_by(artist) %>%
  arrange(desc(popularity)) %>%
  head(n = 16)

top_songs <- top_songs[-c(6), ]

acoustic_plot <- ggplot(data = top_songs, aes(y = reorder(song_title, -acoustic), x = acoustic, fill = artist, 
                           shape = genre)) + geom_col()
acoustic_plot + labs(title = "Acoustic Rating", subtitle = "Top 10 Songs", x = "Acoustic", y = "Song Title") + guides(fill = guide_legend(title = "Artist", title.position = "top"))

```


```{r}

dance_plot <- ggplot(data = top_songs, aes(y = reorder(song_title, -danceability), x = danceability, fill = artist, 
                           shape = genre)) + geom_col()
dance_plot + labs(title = "Danceability Rating", subtitle = "Top 10 Songs", x = "Danceability", y = "Song Title") + guides(fill = guide_legend(title = "Artist", title.position = "top"))


```


```{r Energy Bar Plot, echo=FALSE}

energy_plot <- ggplot(data = top_songs, aes(y = reorder(song_title, -energy), x = energy, fill = artist, 
                           shape = genre)) + geom_col()
energy_plot + labs(title = "Energy Rating", subtitle = "Top 10 Songs", x = "Energy", y = "Song Title") + guides(fill = guide_legend(title = "Artist", title.position = "top"))

```


```{r Loundess Bar Plot, echo=FALSE}

loudness_plot <- ggplot(data = top_songs, aes(y = loudness, x = reorder(song_title, -loudness), fill = artist, 
                           shape = genre)) + geom_col()
loudness_plot + labs(title = "Loudness Rating", subtitle = "Top 10 Songs", x = "Song Title", y = "Loudness") + guides(fill = guide_legend(title = "Artist", title.position = "top")) + theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

```{r Speechiness Bar Plot, echo=FALSE}

speechiness_plot <- ggplot(data = top_songs, aes(y = reorder(song_title, -speechiness), x = speechiness, fill = artist, 
                           shape = genre)) + geom_col()
speechiness_plot + labs(title = "Speechiness Rating", subtitle = "Top 10 Songs", x = "Speechiness", y = "Song Title") + guides(fill = guide_legend(title = "Artist", title.position = "top"))

```


```{r Tempo Bar Plot, echo=FALSE}

tempo_plot <- ggplot(data = top_songs, aes(y = reorder(song_title, -tempo), x = tempo, fill = artist, 
                           shape = genre)) + geom_col()
tempo_plot + labs(title = "Tempo Rating", subtitle = "Top 10 Songs", x = "Tempo", y = "Song Title") + guides(fill = guide_legend(title = "Artist", title.position = "top"))

```

```{r Key Bar Plot, echo=FALSE}

key_plot <- ggplot(data = top_songs, aes(y = reorder(song_title, -key), x = key, fill = artist, 
                           shape = genre)) + geom_col() + scale_x_continuous(breaks = seq(0, 11, 1), limits = c(0, 11))
key_plot + labs(title = "Key", subtitle = "Top 10 Songs", x = "Key", y = "Song Title") + guides(fill = guide_legend(title = "Artist", title.position = "top"))

```

```{r}

acoustic_plot <- ggplot(data = top_songs, aes(y = reorder(song_title, -liveness), x = liveness, fill = artist, 
                           shape = genre)) + geom_col()
acoustic_plot + labs(title = "Liveness Rating", subtitle = "Top 10 Songs", x = "Liveness", y = "Song Title") + guides(fill = guide_legend(title = "Artist", title.position = "top"))

```

\hfill\break 

# Hierarchical Models

## The Complete Pooled Model
The assumption for this section is that we have a response variable $Y$, but no predictors $X$. A few variables were selected and artist levels were reordered according to their mean song popularity. The **Artist Means** data frame summarizes the number of songs from each artist and their mean popularity.

\hfill\break 

```{r Load Hierarchical Data}

# Load data
spotify <- read.csv("all_tracks_updated.csv", header = TRUE, sep = ",",
                    na.strings = c("", " ", "NA"))

# Drop "Various Artists" 
drop <- "Various Artists"
spotify <- spotify[!spotify$artist == drop, ]

# Drop genre "Other"
drop <- "Other"
spotify <- spotify[!spotify$artist == drop, ]

# Drop artists where popularity is 0
drop <- 10
spotify <- spotify[!spotify$popularity <= drop, ]


```

```{r Selecting data to focus, echo=FALSE}

spotify_mutated <- spotify %>% 
  select(artist, name, popularity) %>%
  mutate(artist = fct_reorder(artist, popularity, .fun = "mean"))

# Number of songs
nrow(spotify_mutated)

# Number of artists
nlevels(spotify_mutated$artist)

```

```{r}

# Group by artist and summarize popularity
artist_means <- spotify_mutated %>% 
  group_by(artist) %>%
  summarize(count = n(), popularity = mean(popularity))

artist_means %>%
  slice(1:2, 372:373)

```

```{r Density Plot}

# Create density plot of popularity across all songs
# Variability in popularity from song to song
ggplot(spotify_mutated, aes(x = popularity)) +
  geom_density() + 
  labs(title = "Density Plot Across all Songs", x = "Popularity", y = "Density")

```

### Intercept only model with no predictors
* By substituting $\beta_0$ for global mean $\mu$, this becomes an intercept-only regression model with no predictors $X$. To create a complete pooled model, **stan_glm()** with the following formula **popularity ~ 1** where **1** means "intercept-only" is used.

```{r Complete Pooled Model, message=FALSE, warning=FALSE}

# Prior intercept "normal" family 
# location = prior mean (60) & scale defaults to 2.5
# if autoscale is set to TRUE, scales will be further adjusted and regression 
# coefficients may be additionally modified internally 
complete_pooled <- stan_glm(
  popularity ~ 1,
  data = spotify_mutated, family = gaussian, 
  prior_intercept = normal(60, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000 * 2, seed = 84735
)

prior_summary(complete_pooled)

```


```{r Complete Pooled Summary}

complete_pooled_summary<- tidy(complete_pooled, 
                            effects = c("fixed", "aux"),
                       conf.int = TRUE, conf.level = 0.80)

kable(complete_pooled_summary, 
      caption = "Complete Pooled Summary") %>%
  kable_styling(bootstrap_options = c("bordered", "striped", "responsive")) %>%
  column_spec(1, width = "5em")

```


```{r}

set.seed(84735)
predictions_complete_pooled <- posterior_predict(complete_pooled, 
                                                 newdata = artist_means)

ppc_intervals(artist_means$popularity, 
              yrep = predictions_complete_pooled,
              prob_outer = 0.80) +
  ggplot2::scale_x_continuous(breaks = c(0, 50, 100, 150, 200, 
                                         250, 300, 350, 400)) +
  xlab("Level (Artist)") + ylab("Popularity Rating")

```

## The No Pooled Model
* A no pooled model considers each artist and separately analyzes their popularity. Though there are no predictors, in this section, **artist** will be treated as a predictor. This will limit the understanding to only the artists in this data set, but will help understand this model. A density plot is shown below to begin exploration.

```{r Density Plot by Artist}

ggplot(spotify, aes(x = popularity, group = artist)) +
  geom_density() + labs(title = "Density Plot of Popularity by Artist")

```

```{r No Pooled Model, message=FALSE, warning=FALSE}

no_pooled_model <- stan_glm(
  popularity ~ artist - 1,
  data = spotify, family = gaussian,
  prior = normal(60, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 2000 * 2, seed = 84735
)


```

```{r No Pooled Model Posterior Prediction}

set.seed(84735)
predictions_no_pooled <- posterior_predict(no_pooled_model, 
                                        newdata = artist_means)

ppc_intervals(artist_means$popularity, yrep = predictions_no_pooled,
              prob_outer = 0.80) +
  ggplot2::scale_x_continuous(breaks = c(0, 50, 100, 150, 200, 250, 300,
                                         350, 400)) +
  xlab("Level (Artist)") + ylab("Popularity Rating")


```

## The Partial Pooled Model 

### The model contains three layers: 
* Layer 1: how song popularity varies WITHIN artist $j$
* Layer 2: how popularity $\mu_j$ varies BETWEEN artists
* Layer 3: prior understanding of the entire Spotify population (global parameters $\mu, \sigma_y, \sigma_{\mu}$)

The **stan_glmer()** function with formula **popularity ~ (1 | artist)** is used to indicate that **artist** is a grouping variable instead of being a predictor of popularity. For the posterior simulation the MCMC (Markov chain Monte Carlo) method will be used. The MCMC method is used to approximate the posterior distribution of a parameter of interest by random sampling in a probabilistic space. 

```{r Partial Pooled Model, message=FALSE, warning=FALSE}

spotify_partial_pooled<- stan_glmer(
  popularity ~ (1 | artist),
  data = spotify, family = gaussian,
  prior_intercept = normal(60, 2.5, autoscale = TRUE),
  prior_aux = exponential(1, autoscale = TRUE),
  prior_covariance = decov(reg = 1, conc = 1, shape = 1, scale = 1), 
  chains = 4, iter = 5000 * 2, seed = 84735
)

priors <- prior_summary(spotify_partial_pooled)
priors$prior_aux

```

```{r MCMC, message=FALSE, warning=FALSE, include=FALSE}

mcmc_trace(spotify_partial_pooled)
mcmc_dens_overlay(spotify_partial_pooled)
mcmc_acf(spotify_partial_pooled)
neff_ratio(spotify_partial_pooled)
rhat(spotify_partial_pooled)

```

```{r PP_Check}

pp_check(spotify_partial_pooled) +
  xlab("Popularity")

```

```{r PP Data Frame, include=FALSE}

spotify_pp_df <- as.data.frame(spotify_partial_pooled)

spotify_pp_df %>% 
  colnames() %>% 
  as.data.frame() %>% 
  slice(1:3, 373:376)


```

### Global parameters analysis
* The parameters $\mu$, $\sigma_y$, and $\sigma_{\mu}$, also known as the global parameters, are shared by all artists within and beyond this particular sample. The parameters and their meaning within the results of **stan_glmer()** output are listed below:  
* $\mu$ = intercept
* $\sigma_y$ = sigma
* $\sigma_{\mu}^2$ = Sigma[artist:(Intercept), (Intercept)]

A summary of the partial pooled models is shown below. The "fixed" effect term signifies "non-varying" or "global": 
```{r Tidy Summary}

set.seed(84735)
tidy(spotify_partial_pooled, effects = "fixed", 
     conf.int = TRUE, conf.level = 0.80)

```

```{r}

set.seed(84735)
tidy(spotify_partial_pooled, effects = "ran_pars")

```

The variability in song popularity between artists and differences in popularity of the songs within an artist is calculated: 
```{r Variability in song popualarity between and within artists}

# Variability of song popularity BETWEEN artists
8.22^2 / (8.22^2 + 7.57^2) * 100

# Variability of song popularity WITHIN artists
7.57^2 / (8.22^2 + 7.57^2) * 100

```

From the above calculations, it can be concluded that about **54%** of the variability in song popularity is explained between artists and about **45%** is explained by differences in popularity within an artist. 

```{r Artist Difference Summary}

set.seed(84735)
artist_difference_summary <- tidy(spotify_partial_pooled, effects = "ran_vals",
                                  conf.int = TRUE, conf.level = .80)

artist_difference_summary %>% 
  select(level, conf.low, conf.high) %>%
  arrange(desc(conf.low)) %>%
  slice(1:2, 372:373)

```

```{r Artist Chain Summary, message=FALSE, warning=FALSE}

# MCMC chains for each mu_j
artist_chains <- spotify_partial_pooled %>%
  spread_draws(`(Intercept)`, b[, artist]) %>%
  mutate(mu_j = `(Intercept)` + b)

artist_chains %>%
  select(artist, `(Intercept)`, b, mu_j) %>%
  head(4)

```


```{r Mean Popularity Posterior Summaires, message=FALSE, warning=FALSE}

# Summaries for mu_j
artist_summary_scaled <- artist_chains %>%
  select(-`(Intercept)`, -b) %>%
  mean_qi(.width = 0.80) %>%
  mutate(artist = fct_reorder(artist, mu_j))

artist_summary_scaled %>%
  select(artist, mu_j, .lower, .upper) %>%
  arrange(mu_j) %>%
  tail(30)

```


```{r Artist Summary Scaled Subset}

ggplot(subset(artist_summary_scaled, mu_j >= 78),
       aes(x = artist, y = mu_j, ymin = .lower, ymax = .upper)) +
  geom_pointrange() +
  xaxis_text(angle = 90, hjust = 1)

```

# Posterior Prediction

```{r}

set.seed(84735)
lipa_chains <- spotify_pp_df %>%
  rename(b = `b[(Intercept) artist:Dua_Lipa]`) %>%
  select(`(Intercept)`, b, sigma) %>%
  mutate(mu_lipa = `(Intercept)` + b,
         y_lipa = rnorm(20000, mean = mu_lipa, sd = sigma))

lipa_chains %>%
  mean_qi(y_lipa, .width = 0.80)

artist_summary_scaled %>%
  filter(artist == "artist:Dua_Lipa")

```

```{r Artist Prediction}

set.seed(84735)
jay_sean_chains <- spotify_pp_df %>%
  mutate(sigma_mu = sqrt(`Sigma[artist:(Intercept),(Intercept)]`),
         mu_sean = rnorm(20000, `(Intercept)`, sigma_mu),
         y_sean = rnorm(20000, mu_sean, sigma))

jay_sean_chains %>%
  mean_qi(y_sean, .width = 0.80)

```

```{r Prediction Shortcut, message=FALSE, warning=FALSE}

set.seed(84735)
prediction_shortcut <- posterior_predict(spotify_partial_pooled,
                                         newdata = data.frame(artist = c(
                                           "Dua Lipa", "Jay Sean")))

mcmc_areas(prediction_shortcut, prob = 0.80) +
  ggplot2::scale_y_discrete(labels = c("Dua Lipa", "Jay Sean"))


```